---
title: "missing_value"
output: html_document
date: "2024-02-11"
---


Import library
```{r}
rm(list=ls())
library(dplyr)       
library(tidyr)      
library(readr)      
library(lubridate)
library(arm)
library(tidyselect)
library(ggplot2)
```

Load the data
```{r}
rw.data<-read.csv("data/RWNS_final.csv", header=TRUE,stringsAsFactors=TRUE)
summary(rw.data)
```

#Continuous predictors with missing values

IDACI_n is the only continuous predictor with missing values

```{r}
#In terms of continuous data, we remove the entire row if there is a missing value
new.rw.data <- subset(rw.data, !is.na(IDACI_n))
new.rw.data$homework[new.rw.data$homework == "none"] <- "missing"
```


#Categorical predictors with missing values

Firstly I calculate the percentage of missing data among the categorical predictors(x/13448)

hiquamum <- 6.80% of the data：similar as “no_qualification”, can be removed

*homework* <- 9.87% need to combine “missing” and “none” together (this two are the same)

attitude <- 7.42%: similar as “ver_low”, can be removed

truancy <- 6.23%: missing value is similar as yes, you can remove them, check the boxplot

*absent* <- 11.4% keep the missing value since the number of missing value here is larger than the number of value 'yes'

*exclude* <- 10.47% keep the missing value since the number of missing value here is larger than the number of value 'yes'

##Method1. Remove the missing value

When the percentage of missing value is small(<2.00%), we remove the missing value
```{r}
new.rw.data <- subset(new.rw.data, (SECshort != "missing" & #0.18% of the data
                                    tuition != "missing" &  #1.60% of the data
                                    fsm !="missing" &       #0.13% of the data
                                    sen != "missing" &      #0.13% of the data
                                    computer != "missing" & #1.60% of the data
                                    parasp != "missing" &   #1.60% of the data
                                    singlepar != "missing"& #0.82% of the data
                                    FSMband != "missing"&   #0.13% of the data
                                    hiquamum != "missing" & #6.80% of the data
                                    attitude != "missing" & #7.42% of the data
                                    truancy != "missing"    #6.23% of the data
                                    ))
```

We would add these values back if the predictor is significant in the model

##Method2 Transform the missing value into a level of the categorical predictor

When the amount of missing data for a categorical value is large, we transform the NAs into missing and treat them as a level of the categorical predictor for further analysis


We have explored that the performance of the model shows similar patterns no matter using k3ma, k3sc or k3en. We would use k3ma for the following analysis.

```{r}
new.rw.data$exclude <- relevel(new.rw.data$exclude, ref = "No")
ks4score.lm <- lm(ks4score ~ k3ma+exclude, data = new.rw.data)
display(ks4score.lm)
```
In the fixed level of k3ma score, missing value gives a slightly lower ks4score than those who have not been excluded from school.

Students who have been excluded from school's ks4score is much lower than those who have not been excluded from school and the missing value.


```{r}
ks4score.lm <- lm(ks4score ~ k3ma+hiquamum, data = new.rw.data)
display(ks4score.lm)
```
Missing value in hiquamum is estimated to have higher ks4score than those with other qualificaitons.


```{r}
ks4score.lm <- lm(ks4score ~ k3ma+homework, data = new.rw.data)
display(ks4score.lm)
```
For a fixed value of k3ma a school with missing value in homework has a lower ks4score than those with or without homework


```{r}
new.rw.data$absent <- relevel(new.rw.data$absent, ref = "No")
ks4score.lm <- lm(ks4score ~ k3ma+absent, data = new.rw.data)
display(ks4score.lm)
```
For a fixed value of k3ma, students with missing value in absent have a higher ks4score than those who have and have not been absent from school.


```{r}
new.rw.data$truancy <- relevel(new.rw.data$truancy, ref = "No")
ks4score.lm <- lm(ks4score ~ k3ma+truancy, data = new.rw.data)
display(ks4score.lm)
```
For a fixed value of k3ma, students with missing value in truancy have a higher ks4score than those who have been truant from school but lower than those who have not been truant from school.

```{r}
ks4score.lm <- lm(ks4score ~ k3ma+attitude, data = new.rw.data)
display(ks4score.lm)
```
For a fixed value of k3ma, students with missing value in attitude have a lower ks4score than those with very_high attitude but higher than those with high attitude.

```{r}
write.csv(new.rw.data, file = "RWNS_cleaned.csv")
```

