---
title: "ST211mini project"
author: "Group4"
output: html
---

## Set up
```{r}
library(dplyr)       
library(tidyr)      
library(readr)      
library(lubridate)
library(arm)
library(tidyselect)
library(ggplot2)
```

# Data Loading
```{r}
data<-read.csv("data/RWNS_cleaned.csv", header=TRUE,stringsAsFactors=TRUE)
```

# Convert variables to factors
```{r}
data$sen <- as.factor(data$sen)
data$computer <- as.factor(data$computer)
data$gender <- as.factor(data$gender)
```

# Build the initial regression model
```{r}
model <- lm(ks4score ~ . - fiveem - sen - computer - gender - k3ma - k3sc - singlepar - tuition - exclude - absent, data = data)

## have a look at the model
summary(model)
```

# Summarize the initial model
```{r}
par(mfrow=c(2,2))
plot(model, which=c(1,2))
hist(rstandard(model), freq = FALSE ,
     main="Histogram of standardised residuals",
     cex.main=0.8, xlab="Standardised residuals")
```

Our initial diagnostic plot shows that the variance of the residuals may not be constant (possible heteroscedasticity). Furthermore, the normal Q-Q plot and histogram shows that while the residuals are mostly normally distributed, there are some deviations in the tails, which may indicate a potential outlier or extreme value that could impact the regression analysis.

# Second version of the model 
```{r}
model.1 <- lm(ks4score ~ . - fiveem - sen - computer - gender - k3ma - k3sc - singlepar - tuition - exclude - absent - fsm - house - IDACI_n, data = data)

## have a look at the model
summary(model.1)
```

# Remove outliers
```{r}
show_outliers<-function(the.linear.model,topN){
  #length of data
  n=length(fitted(the.linear.model))
  #number of parameters estimated
  p=length(coef(the.linear.model))
  #standardised residuals over 3
  res.out<-which(abs(rstandard(the.linear.model))>3) #sometimes >2
  #topN values
  res.top<-head(rev(sort(abs(rstandard(the.linear.model)))),topN)
  #high leverage values
  lev.out<-which(lm.influence(the.linear.model)$hat>2*p/n)
  #topN values
  lev.top<-head(rev(sort(lm.influence(the.linear.model)$hat)),topN)
  #high diffits
  dffits.out<-which(dffits(the.linear.model)>2*sqrt(p/n))
  #topN values
  dffits.top<-head(rev(sort(dffits(the.linear.model))),topN)
  #Cook's over 1
  cooks.out<-which(cooks.distance(the.linear.model)>1)
  #topN cooks
  cooks.top<-head(rev(sort(cooks.distance(the.linear.model))),topN)
  #Create a list with the statistics -- cant do a data frame as different lengths 
  list.of.stats<-list(Std.res=res.out,Std.res.top=res.top, Leverage=lev.out, Leverage.top=lev.top, DFFITS=dffits.out, DFFITS.top=dffits.top, Cooks=cooks.out,Cooks.top=cooks.top)
  #return the statistics
  list.of.stats}
```

```{r}
#Get outliers
outliers <- show_outliers(model.1, topN = 5) 

# Combine all outlier indices into a single vector
all_outliers <- unique(c(outliers$Std.res, outliers$Leverage, outliers$DFFITS, outliers$Cooks))

# Remove outliers from the dataset
cleaned_data <- model.1$model[-all_outliers, ]

# Refit the model without outliers
clean_model <- lm(ks4score ~ . - fiveem - sen - computer - gender - k3ma - k3sc - singlepar - tuition - exclude - absent - fsm - house - IDACI_n, data = data)

# Summarize the new model
summary(clean_model)
```
# Summarize the final model
```{r}
par(mfrow=c(2,2))
plot(clean_model, which=c(1,2))
hist(rstandard(model), freq = FALSE ,
     main="Histogram of standardised residuals",
     cex.main=0.8, xlab="Standardised residuals")
```

